Towards an Ecological Trait-data Standard

### 2018-09-13 manuscript/manuscript.Rmd
update manuscript and biblio for internal review

1. Trait-based approaches are widespread throughout ecological research, offering great potential for trait data to deliver general and mechanistic conclusions. Accordingly, a wealth of trait data is available for many organism groups, but, due to a lack of standardisation, these data come in heterogeneous formats.
2. We review current initiatives and infrastructures for standardising trait data and discuss the importance of standardisation for trait data hosted in distributed open-access repositories.
3. In order to facilitate the standardisation and harmonisation of distributed trait datasets, we propose a standardised vocabulary to be used for storing and sharing ecological trait data. 
4. This vocabulary allows trait datasets from heterogeneous sources to be merged more easily into databases, and facilitates the creation of data interfaces between software tools for trait-data handling and analysis. 
5. By enabling decentralised trait-data standardisation, we intend to facilitate the exchange and analysis of trait data for a broader ecological research community and enable global syntheses of traits across a wide range of taxa and ecosystems.

@@

Functional traits are phenotypic  (i.e. morphological, physiological, behavioral) characteristics that are related to the fitness and performance of an organism [@mcgill06; @violle07]. Because trait-based approaches allow studying both patterns and mechanisms [@lavorel02; @diaz16], recent years have seen a proliferation of trait-based research in a wide range of fields.
Trait-based studies have been conducted in a wide range of thematic areas ranging from the evolutionary basis of individual-level properties [@salguero-gomez16] to global patterns of biodiversity [@diaz16] and ecosystem functioning [@allan15; @bello10]. 
The trait framework relates losses of ecosystem function to changes in  the functional composition of species assemblages [@mouillot13; @perovic15]. This offers the mechanistic background to relate biodiversity to climate change or local anthropogenic land use [@diaz11; @lavorel12; @allan15]. <!-- ;@wood15;  @diaz11; @jennings09; @hopfenmuller14; @mouillot13;  @vandewalle10; @perovic15-->
Using traits is also a promising means of bypassing taxonomic impediment, i.e. the fact that a majority of species are yet undescribed and little is known of their interactions with the environment and other organisms. This is because functional traits allow us to infer the ecological role of organisms  from their apparent features, regardless of their taxonomic identity [@leprovost17; @duarte11; @schrodt15].

<!-- Comparative studies of traits inform researchers about physiological and evolutionary trade-offs, and help identifying economic strategies of organisms beyond taxonomy [@menezes10; @forister15;  @siefert15; @gross09; @wright04]. Inferring an unknown functional role or ecological function of individuals from their apparent features [@duarte11; @schrodt15] is a promising venue to bypass taxonomic impediment, i.e. the fact that a majority of species are yet undescribed and little is known of their interactions with the environment and other organisms. 
Mutual 'trait-matching' of co-evolved organisms enables the inference of ecological interactions from observed features of live specimens [ @kalinkat13; @dehling14a] which may be used to inform ecological networks [@weiner14; @miranda13; @digel11; @schleuning15] and parametrise ecological population-dynamic models [@schneider12; @kalinkat13]. These analyses are also a promising venue for a multitrophic community ecology aiming to unravel dependencies and synchrony in trait spaces between interacting functional groups [@soliveres16]. <!-- GLP: Maybe not the best example as they are not looking at trait diversty and species interactions ? 
Put instead:
Gravel et al. 2016 The meaning of functional trait composition of food webs for ecosystem functioning
And Coux et al. 2016 Linking species functional roles to their network roles
?
-->
<!--
Since traits are so closely related to ecological mechanisms, they have proven useful for scaling individual level physiology and behaviour to functional diversity  of communities [@laliberte10a; @petchey09; @villeger08a; @hillerislambers12;@diaz98; @lavorel02], processes at the ecosystem scale [@fontaine05; @simons16; @gossner16; @moretti13;  @kuhsel15;  @birkhofer14], as well as macro-patterns of biodiversity at the global scale [@diaz16; e.g. @reich97]. <!-- Traits can be used to predict extinction risk [@fritz10; @pearson14] and susceptibility to environmental toxins [@rubach11].--> <!-- GLP: What do you mean by processes ? I’m not sure that the literature cited is always appropriate here ?
-->
<!--
Most importantly, the trait framework relates key ecosystem services for human well-being [@diaz11; @lavorel12; @allan15] back to changes in biodiversity in consequence to global change or local anthropogenic land use [ @diaz07; @gossner16; @mangels17; @jennings09; @simons16;  @mouillot13;  @vandewalle10;  @birkhofer15;@gamez-virues15; @birkhofer17]. 
--> 
<!-- In recent years, trait-based ecology has thus grown into a major field of ecological research, which continues to integrate a rapidly increasing number of fundamental and applied research questions. 
Given the expected future challenges in dealing with climate change and managing land use,  global trait synthesis is an interdisciplinary, cross-topic field with great potential. -->

<!-- 2. assemble trait data into databases -->

Many issues in trait-based research arise when compiling datasets from multiple sources. Data may differ in taxonomic nomenclature and resolution (e.g. reported on species level or aggregated on higher taxonomical orders), the scale and place of the study context, or the accurracy of the methodology applied in measurements. These differences are not always documented in the metadata accompanying a dataset. 
All of these factors render trait data extremely heterogeneous and make the task of data compilation time-consuming or even prohibitive.
However, fully exploiting the potential of trait-based approaches relies heavily on the broad availability and compatibility of trait data to achieve sufficient taxonomic and regional coverage, both of present-day taxa ase well as in evolutionary deep-time.

<!-- 3. trait databases have been developed, with standard structure and terms --> 
To this end, the number of available trait datasets is increasing rapidly. 
In the past, trait data have been standardised and compiled in centralised databases for specific organism groups and regional scope, often centred around particular research questions [e.g. PanTHERIA, @jones09; TRY, @kattge11; AmphiBio, @oliveira17]. 
<!--These databases are then published as static data tables (i.e. product of a one-time effort as it is the case of AmphiBIO), or in the form of sophisticated online portals, which attract data submissions from a focal community of researchers and provide research networks for consensus building in data resolution and measurement methods (e.g. TRY, www.try-db.org). Furthermore,--> These initiatives map heterogeneous data into a common scheme and, importantly, may also offer access control and data usage policies. As such, they protect the rights of the original data providers while simplifying data queries for synthesis researchers. 
Besides initiatives aiming at assembling data, other tools to enable the compatibility of data across databases are being developed. These include semantic-web standards [@wieczorek12; @page08] and ontologies of standard terms [@walls12; @garnier17].
<!-- Solving the numerous issues of data assemblage and harmonisation in centralised infrastructures requires a huge effort in costs and personnel. -->
<!-- 4. distributed trait data must be integrated -->

Meanwhile, open-science reaches the mainstream: it has become the declared goal of an open biodiversity knowledge management (http://www.bouchoutdeclaration.org/) and is increasingly demanded by journals and public research funding [@emerson15; @allianceofgermanscienceorganisations10; @royalsocietysciencepolicycentre12; @swan12; @allison15]. As a result, an increasing number of individual research projects publish their primary data on file hosting services like Figshare.com, Dryad (datadryad.org), Researchgate.net, or Zenodo.org, where no data standards are forced upon the uploaded material. <!-- The data standards mentioned above, that have been proposed for relational trait-databases, do hardly apply for these individual datasets.-->
It is likely that trait data will become increasingly available, but a lack of data and metadata standardisation will hamper the efficient re-use and synthesis of published datasets.

In this paper, we review existing trait databases and online portals, as well as initiatives for standardisation. We discuss current practice and the importance of data standards for trait-based research, and we identify current deficits in standardisation from a pragmatic view of data providers and data users. Based on these considerations, we propose a versatile vocabulary for describing trait datasets, that builds upon and is compatible with existing terminology standards for biodiversity data. 
Since this standard vocabulary relies on the adoption of a broad research community we discuss incentives for its use and lay out mechanisms for a community-based future development towards an accessible ecological trait data standard.
<!--Finally, we present an R package that assists the harmonisation of trait data from distributed sources. -->

<!-- FDS: Why 'ecological' trait data? what is special about this subject? Environmental context, community context, space and time, attributes of functional role in the ecosystem and energetic demands, interactions (not covered right now?), correlation and economic trade-offs in trait evolution -->

<!-- PM: Can I suggest a glossary of terms – this would be really helpful for people who are not data management specialists

FDS: see Garnier et al. but may contain: 
-->

# A review of initiatives for trait-data standardisation

In this section, we review four types of initiatives that are of relevance for trait-data standardisation (see Glossary in Table 1 for italicised terms):

@@

For trait data, there are typical issues arising from the variability of data structures. 
For instance, the column descriptions and terminology applied to taxa and traits are mostly project specific, and rarely chosen to allow translation into larger database initiatives. Furthermore, metadata varies in its detail, e.g. for documenting descriptions of variables, measurement procedures or sampling context [@kattge11a]. 
In terms of structure, trait data usually are reported in a species$\times$traits wide-table format. In this format, each row contains a species (or taxon) for which multiple traits are reported in columns. 
Similarly, when reporting raw data, researchers place observations of individual organisms in rows with multiple trait measurements applied to the same individual across multiple columns. 
Variability in the number and meaning of columns in these *data tables* requires tedious manual adjustments when merging multiple datasets [@wickham14].

@@

A global overview of existing trait data for all taxa and trait types is difficult to obtain. Therefore, in an attempt to collate a list of existing distributed datasets, we initiated a living spreadsheet (https://goo.gl/QxzfHy) which lists published trait datasets, their regional and taxonomic focus, the number and scope of traits covered, their location on the internet and the terms of use (see Appendix A for an excerpt of this list). We invite data owners and users to add further trait datasets to this spreadsheet.  
As it stands, the decentralisation and the lack of data standardisation of low-threshold online repositories renders the compilation of data into larger collections inefficient and reduces the potential of many published datasets to be re-used and combined into broad synthesis analysis.

@@

<!-- community database backend, but static publication --> 
In the past two decades, many distributed trait datasets have been aggregated and harmonised into greater collections with particular taxonomic or regional focus [e.g. @klotz02; @iversen17; @kleyer08; @myhrvold15; @kissling14; @jones09; @oliveira17, see Appendix A table A1].<!--[@wright04; ] -->
While mostly concerned with issues of heterogeneity in units or factor levels, and aiming for high taxonomic coverage, few of these datasets apply a standardised terminology for taxa or traits that would allow them to be efficiently related to other databases.<!--and aiming for taxonomic completeness, the products of these initiatives employ a project specific structure and vocabulary. -->
Documentation of metadata and methodology differs in the level of detail, depending on the research focus of the initiative. 
Just as the individual datasets described above, many of these databases are published as static data tables on journal websites or open-access file hosting platforms and may be updated irregularly.

As they deal with much larger amounts of data, initiatives that form around natural history museum collections are traditionally more concerned with standardisation. 
Concerning organism traits, with the digitisation efforts that are currently undertaken in many museum collections [@vollmar10; @blagoderov12], supported by citizen science crowd-sourcing (e.g. www.markmybird.org), data on body measurements are likely to grow exponentially in the near future. 
For example, the VertNet database compiled and harmonised large quantities of vertebrate trait data with the aim of mobilising measurements from collections [@guralnick16]. The resulting data are published as versioned data tables which are updated as new data sources become available (vertnet.org).

@@

Microbial ecologists also make frequent use of trait-based approaches to assess genomic function and describe functional diversity at the community level [@krause14; @fierer14;  @fierer12]. Here, 'operational taxonomic units' (OTUs) are derived from metagenomic analysis [@torsvik02; @langille13]. Databases are also used to interpret OTUs in terms of their functional role [e.g. the KEGG orthology, @kanehisa12]. 
For animals, a single unified platform and harmonising scheme for animal trait data is still lacking. The reason for this may be that harmonising trait data on animals, which span multiple trophic levels and possess diverse body plans, is a more complex task than for plants [@moretti17]. Nonetheless, initiatives for particular groups of animals do exist, such as the BETSI database collects traits on soil invertebrates [http://betsi.cesab.org/; @pey14], and the Carabids.org web portal collects traits of carabid beetles (http://www.carabids.org/).

The platforms incentivise data submissions by offering increased data visibility and usage, while providing data use policies that secure author attribution and potentially co-authorship. 
However, the upload of data to these centralised databases does currently not fulfill the criteria demanded by journals and funding agencies for primary data publication with long-term availability, citable DOIs and no access restrictions. With data being publicly available, the role of database initiatives in standardising data and making them more accessible is paramount.

@@

By filling this gap, data-brokering services [e.g. the German Federation for Biological Data; gfbio.org; @diepenbroek14] or data management systems for scientific projects (e.g. KNB and its open-source database back-end Metacat, https://knb.ecoinformatics.org/; Diversity Workbench, www.diversityworkbench.net; BExIS, http://bexis2.uni-jena.de/; ) are likely to gain importance. These services simplify and direct the standardised upload of research data and descriptive metadata into reliable and interlinked data infrastructures. One goal of such initiatives is to facilitate data usage in publications and provide standardisation for researchers, for instance by providing terminologies and ontologies for biodiversity data, and by consulting on publication licenses. Another solution for data-users to access trait data in a structured way is offered by FENNEC [@ankenbrand18], an online tool or self-hosted service to extract trait information from multiple sources for a target species community.  <!-- AG:  EDIT Platform for Cybertaxonomy (cybertaxonomy); NEON Dataportal http://data.neonscience.org/home;   -->

## Conclusion of review

Initiatives for standardisation (e.g. ontologies and data standards) and platforms for data management (e.g. database and data management platforms) provide great visibility and improve interconnectedness of datasets, but raise relatively high thresholds for data and metadata preparation. 
On the other end, public repositories on file-hosting services offer the publication of scientific primary data according to current funding and journal criteria, attracting a wealth of heterogeneous trait datasets. Due to the low thresholds concerning data standardisation, compilation of these distributed data sets is currently laborious. 
The goal must be to better integrate these distributed data into the global biodiversity data-management ecosystem by creating awareness for data standardisation on the side of data providers. 
We propose the development of tools and vocabularies that impose low thresholds and offer high pay-off in the visibility and interconnectedness of published data.

@@

As a response to the challenges outlined above, we propose a versatile vocabulary for trait-based ecological research.
The aim of the vocabulary is to cover the variety of trait-based approaches and their different degrees of measurement detail. 
Rather than describing a data structure for relational databases, the vocabulary is intended as a more inclusive terminology, that can be used in three major use cases:

@@

By using this standard vocabulary, authors can ensure that the description of trait measurements that are uploaded to distributed data repositories will be unambiguous and generally applicable. It will facilitate re-use of data for future data aggregation initiatives and data synthesis and ensure long-term accessibility.

In designing this vocabulary, we drew on the combined expertise of empirical biodiversity researchers (data providers), biodiversity synthesis researchers (data users), and biodiversity informatics researchers (data managers).
We paid particular consideration to existing data standards (particularly @kattge11; @kattge11a; @garnier17; @parr16) to ensure compatibility of our proposed data structure with major trait databases and existing standards for biodiversity data management. 
Here, the use of identifiers ('IDs') for the individual measurement observations ('measurementID'), specimens ('occurrenceID'), sampling events ('eventID'), or taxa ('taxonID') is key to map two-dimensional data onto the structure of relational databases. 
<!--Besides being used for the publication of datasets, the standard vocabulary could be imposed in webservices or download tools, e.g. APIs that provide direct access to online databases. -->
The vocabulary proposed is intended to form the foundations of a standard exchange format that can be expanded and revised by the wider community of researchers using trait-based approaches in ecology.
<!--
## Why use the vocabulary

@@

## Terms of the standard vocabulary

The standard vocabulary is accessible at https://ecologicaltraitdata.github.io/ETS/.
The core terms describe a dataset according to the Entity-Quality model described above [@garnier17]: each entry describes a trait value (i.e. quality) observed on an individual or population (i.e. entity), of a biological taxon.
Beyond these core observations, further information might be available that are related to the taxonomic assignment, or that put the reported fact, measurement or sampling event in a broader observation context (including geolocation and date information). 
These information may be useful for future analysis of the causal reasons of trait variation and should always be published along with the core data. For this case, we offer three extensions of the core vocabulary ("Taxon", "Measurement or Fact", and "Occurrence"), which may simply be added as extra columns to the core dataset (see below for best practice when publishing primary data).
Additional terms are provided for metadata and for relating trait names to definitions and external ontologies or thesauri (see section on metadata below).

Some of the terms expand and refine terms of the Darwin Core Extensions (http://rs.tdwg.org/dwc/terms/) or Glossary of EOL (http://eol.org/info/516). Wherever our glossary refines or duplicates existing terms, we indicate this in the fields 'refines' or 'identical', respectively.
Any further columns on co-variates or measurement context, or other dataset-specific content, might draw from terms provided by Darwin Core and it's extensions, or be added as user-defined columns. However, these should be clearly specified in the meta-information accompanying the dataset. 
The scope of the vocabulary may not yet cover all aspects of morphological and evolutionary perspectives. 
Also, information about interactions between species are not within the scope of the Entity-Quality Model, but may easily be combined with trait data by using other extensions of DwC.
Therefore, we invite researchers to contribute to the next iterations of the standard vocabulary and develop own applications and ontologies that interact with it.

### Specification of core terms

@@

### Extensions for additional data layers

@@

When publishing primary data, priority is on containing measurement values while minimizing the risk of purporting errors through data conversion or aggregation. Thus, ideally, data are published as they have been recorded, i.e. in original units and raw values, along with standardised values that match a unified and compatible scheme of a data standard. This is true for the core terms, i.e. the trait value (the quality) and taxon name (entity), but also for any co-variates reported along with it. In the case of the core terms, data providers should use the '\*Std'-terms for standardised entries. For co-variates, the extensions of the vocabulary provide column names for standardised values only, e.g. elevation above sea-level. If any conversion was applied to obtain these data, users should keep the original values in user-defined columns, e.g. by appending '\*Orig' to the column name. Also, additional co-variates, e.g. on environmental growth conditions of the specimen, should be provided in user-defined columns. Any user-defined columns must be adequately defined in the accompanying meta-information.

When applying the vocabulary in a primary data publication, most authors will chose a two-dimensional spreadsheet rather than relational databases for reasons of reduced complexity. 
For spreadsheets, it is recommended to use a observation long-table format for the data (Fig. 2 *b*), rather than a species$\times$traits matrix (Fig. 2 *a*). 
As the long-table format draws from a defined set of columns, merging datasets is easier. Long-table datasets also purport multiple advantages for data manipulation [e.g. filtering, sub-setting and aggregating data, @wickham14].

>>>>

<!-- 
The Abstract must not exceed 350 words and should list the main results and conclusions, using simple, factual, numbered statements:

Point 1: set the context for and purpose of the work;

Point 2: indicate the approach and methods;

Point 3: outline the main results;

Point 4: identify the conclusions and the wider implications.
-->

1. Trait-based approaches are widespread throughout ecological research, offering great potential for trait data to deliver general and mechanistic conclusions. Accordingly, a wealth of trait data is available for many organism groups, but, due to a lack of standardisation, these data come in heterogeneous formats.
2. We review current initiatives and infrastructures for standardising trait data and discuss the importance of standardisation for trait data hosted in distributed open-access repositories. 
3. In order to facilitate the standardisation and harmonisation of distributed trait datasets on the data-provider side, we propose a standardised vocabulary to be used for storing and sharing ecological trait data. We discuss impediments and incentives for a wide adoption of such a standard by data-providers. 
4. The use of a standard vocabulary allows trait datasets from heterogeneous sources to be compiled more easily into databases, and facilitates the creation of interfaces between software tools for trait-data handling and analysis. By aiding decentralised trait-data standardisation, our data standard facilitates the use of trait data for a broader ecological research community and enables global syntheses across a wide range of taxa and ecosystems.

@@

Functional traits are phenotypic (i.e. morphological, physiological, behavioral) characteristics that are related to the fitness and performance of an organism [@mcgill06; @violle07]. 
Recent years have seen a proliferation of trait-based research in a wide range of fields. 
Traits bridge the evolutionary basis of individual-level properties [@salguero-gomez16] to global patterns of biodiversity [@diaz16] and relate losses of ecosystem function to changes in  the functional composition of species assemblages [@mouillot13; @perovic15; @bello10]. This offers the mechanistic background for linking biodiversity to climate change or local anthropogenic land use [@diaz11; @lavorel12; @allan15].
 <!-- ;@wood15;  @diaz11; @jennings09; @hopfenmuller14; @mouillot13;  @vandewalle10; @perovic15-->
Using traits is also a promising means of bypassing taxonomic impediment, i.e. the fact that a majority of species are yet undescribed and little is known of their interactions with the environment and other organisms. This is because functional traits allow us to infer the ecological role of organisms  from their apparent features, regardless of their taxonomic identity [@leprovost17; @duarte11; @schrodt15].

<!-- Comparative studies of traits inform researchers about physiological and evolutionary trade-offs, and help identifying economic strategies of organisms beyond taxonomy [@menezes10; @forister15;  @siefert15; @gross09; @wright04]. Inferring an unknown functional role or ecological function of individuals from their apparent features [@duarte11; @schrodt15] is a promising venue to bypass taxonomic impediment, i.e. the fact that a majority of species are yet undescribed and little is known of their interactions with the environment and other organisms. 
Mutual 'trait-matching' of co-evolved organisms enables the inference of ecological interactions from observed features of live specimens [ @kalinkat13; @dehling14a] which may be used to inform ecological networks [@weiner14; @miranda13; @digel11; @schleuning15] and parametrise ecological population-dynamic models [@schneider12; @kalinkat13]. These analyses are also a promising venue for a multitrophic community ecology aiming to unravel dependencies and synchrony in trait spaces between interacting functional groups [@soliveres16]. <!-- GLP: Maybe not the best example as they are not looking at trait diversty and species interactions ? 
Put instead:
Gravel et al. 2016 The meaning of functional trait composition of food webs for ecosystem functioning
And Coux et al. 2016 Linking species functional roles to their network roles
? -->
<!--
Since traits are so closely related to ecological mechanisms, they have proven useful for scaling individual level physiology and behaviour to functional diversity  of communities [@laliberte10a; @petchey09; @villeger08a; @hillerislambers12;@diaz98; @lavorel02], processes at the ecosystem scale [@fontaine05; @simons16; @gossner16; @moretti13;  @kuhsel15;  @birkhofer14], as well as macro-patterns of biodiversity at the global scale [@diaz16; e.g. @reich97]. <!-- Traits can be used to predict extinction risk [@fritz10; @pearson14] and susceptibility to environmental toxins [@rubach11].-->
<!--
Most importantly, the trait framework relates key ecosystem services for human well-being [@diaz11; @lavorel12; @allan15] back to changes in biodiversity in consequence to global change or local anthropogenic land use [ @diaz07; @gossner16; @mangels17; @jennings09; @simons16;  @mouillot13;  @vandewalle10;  @birkhofer15;@gamez-virues15; @birkhofer17]. 
-->

<!-- 2. trait databases have been developed, with standard structure and terms --> 
However, fully exploiting the potential of trait-based approaches relies heavily on the broad availability and compatibility of trait data to achieve sufficient taxonomic and regional coverage, both of present-day taxa ase well as in evolutionary deep-time. 
To enable such functional ecological research, trait data have been harmonized and compiled into centralised databases for specific organism groups and regional scope, often centred around particular research questions [e.g. PanTHERIA, @jones09; TRY, @kattge11; AmphiBio, @oliveira17]. 
The heterogeneity of data from different research contexts <!-- i.e. differences in taxonomic nomenclature and resolution (e.g. reported on species level or aggregated on higher taxonomical orders), the scale and place of the study context, or the accurracy of the methodology applied in measurements--> render trait data extremely heterogeneous and make the task of data compilation time-consuming and error-prone.
Besides initiatives aiming at assembling data, other tools to enable the compatibility of data across databases are being developed. These include software to access trait data from the internet [e.g. @chamberlain17; @ankenbrand18], semantic-web standards [@wieczorek12; @page08] and ontologies of standard terms [@walls12; @garnier17].

<!-- 4. distributed trait data must be integrated --> 
Meanwhile, open science reaches the mainstream: publication of research data without access restrictions, with structured metadata and in accordance to data standards to enable their reuse, has become the declared goal of an open biodiversity knowledge management (http://www.bouchoutdeclaration.org/) and is increasingly demanded by journals and public research funding [@emerson15; @allianceofgermanscienceorganisations10; @royalsocietysciencepolicycentre12; @swan12; @allison15]. As a result, an increasing number of individual research projects publish their primary data on general-purpose file hosting services (e.g. Figshare.com, datadryad.org, Researchgate.net, or Zenodo.org), where no data standards are forced upon the uploaded material [@wilkinson16]. 
It is thus likely that trait data will become increasingly available, but a lack of data and metadata standardisation will hamper the efficient reuse and synthesis of published datasets.

In this paper, we review existing trait databases and online portals, as well as initiatives for data standardisation from a pragmatic view of data providers and data users. We discuss the current efforts for making trait data visible, accessible and reuseable in downstream data analysis, as demanded by the FAIR principles for scientific data [@wilkinson16]. We show how the current deficit in standardisation hampers the implementation of interoperability and reuse of trait data. 
Based on these considerations, we propose a versatile vocabulary for describing trait datasets, that builds upon and is compatible with existing terminology standards for biodiversity data. 
Since this standard vocabulary relies on the adoption of a broad research community we discuss incentives for its use and lay out mechanisms for a community-based future development towards an accessible ecological trait data standard.
<!--Finally, we present an R package that assists the harmonisation of trait data from distributed sources. -->
<!-- FDS: Why 'ecological' trait data? what is special about this subject? Environmental context, community context, space and time, attributes of functional role in the ecosystem and energetic demands, interactions (not covered right now?), correlation and economic trade-offs in trait evolution -->

# Initiatives for trait-data standardisation

We review four types of initiatives that are of relevance for trait-data standardisation (see Glossary in Table 1 for italicised terms):

@@

For trait data, there are typical issues arising from the variability of data structures. 
For instance, the column descriptions and terminology applied to taxa and traits are mostly project specific, and rarely chosen to allow translation into larger database initiatives. Furthermore, metadata varies in its detail, e.g. for documenting descriptions of variables, measurement procedures or sampling context [@kattge11a]. 
In terms of structure, trait data usually are reported in a species$\times$traits wide-table format. In this format, each row contains a species (or taxon) for which multiple traits are reported in columns. 
Similarly, when reporting raw data, researchers place observations of individual organisms in rows with multiple trait measurements applied to the same individual across multiple columns. 
Variability in the number and meaning of columns in these *data tables* requires tedious manual adjustments when merging multiple datasets [@wickham14].
As it stands, the decentralisation and the lack of data standardisation of data in online repositories renders the compilation of data into larger collections inefficient and reduces the potential of many published datasets to be reused and combined into broad synthesis analysis.

@@

<!--A global overview of existing trait data for all taxa and trait types is difficult to obtain. Therefore, in an attempt to collate a list of existing distributed datasets, we initiated a living spreadsheet (https://goo.gl/QxzfHy) which lists published trait datasets, their regional and taxonomic focus, the number and scope of traits covered, their location on the internet and the terms of use (see Appendix A for an excerpt of this list). We invite data owners and users to add further trait datasets to this spreadsheet.  -->
<!-- FDS: this might be a task for community  building and decentralised registries, as discussed on the OpenTraits platform, e.g. FENNEC or MAD package do that, or a registry for earning badges of data quality would achieve that as a by-product. I think we should not aim for solving this registry problem here. -->

@@

<!-- community database backend, but static publication --> 
In the past two decades, many distributed trait datasets have been aggregated and harmonised into greater collections with particular taxonomic or regional focus [e.g. @klotz02; @iversen17; @kleyer08; @myhrvold15; @kissling14; @jones09; @oliveira17, see Supplementary Material A, table A1].<!--[@wright04; ] -->
While mostly concerned with issues of heterogeneity in units or factor levels, and aiming for high taxonomic coverage, few of these datasets apply a standardised terminology for taxa or traits that would allow them to be efficiently related to other databases.<!--and aiming for taxonomic completeness, the products of these initiatives employ a project specific structure and vocabulary. -->
Documentation of metadata and methodology differs in the level of detail, depending on the research focus of the initiative. 
Just as the individual datasets described above, many of these databases are published as static data tables on journal websites or open-access file hosting platforms and may be updated irregularly.

As they deal with much larger amounts of data, initiatives that form around natural history museum collections are traditionally more concerned with standardisation. 
Concerning organism traits, with the digitisation efforts that are currently undertaken in many museum collections [@vollmar10; @blagoderov12], supported by citizen science crowd-sourcing (e.g. www.markmybird.org), data on body measurements are likely to skyrocket in the near future. 
For example, the VertNet database compiled and harmonised large quantities of vertebrate trait data with the aim of mobilising measurements from collections [@guralnick16]. The resulting data are published as versioned data tables which are updated as new data sources become available (vertnet.org).

@@

Microbial ecologists also make frequent use of trait-based approaches to assess genomic function and describe functional diversity at the community level [@krause14; @fierer14;  @fierer12]. Here, 'operational taxonomic units' (OTUs) are derived from metagenomic analysis [@torsvik02; @langille13]. Databases are also used to interpret OTUs in terms of their functional role [e.g. the KEGG orthology, @kanehisa12]. 
For animals, a single unified platform and harmonising scheme for animal trait data is still lacking. The reason for this may be that harmonising trait data on animals, which span multiple trophic levels and possess diverse body plans, is a more complex task than for plants [@moretti17]. 
Nonetheless, initiatives for particular groups of animals do exist: for instance, the BETSI database collects traits on soil invertebrates [http://betsi.cesab.org/; @pey14], the Carabids.org web portal collects traits of carabid beetles (http://www.carabids.org/), or the Coral Trait Database compiles ecological data on coral observations [@madin16].

The role of database initiatives in standardising data and making them more accessible is paramount. 
The trait-data platforms incentivise data submissions by offering increased data visibility and usage, while providing data-use policies that secure author attribution and, potentially, co-authorship. 
However, the upload of data to these centralised databases does often not fulfill all of the criteria that are increasingly demanded by journals and funding agencies for primary data publication with long-term availability, citable DOIs and no access restrictions.

@@

By filling this gap, data-brokering services [e.g. the German Federation for Biological Data; gfbio.org; @diepenbroek14] or data management systems for scientific projects (e.g. KNB and its open-source database back-end Metacat, https://knb.ecoinformatics.org/; Diversity Workbench, www.diversityworkbench.net; BExIS, http://bexis2.uni-jena.de/) are likely to gain importance. These services simplify and direct the standardised upload of research data and descriptive metadata into reliable and interlinked data infrastructures. One goal of such initiatives is to provide standardisation of data and facilitate data usage, for instance by mapping unambiguous terminologies and ontologies for biodiversity data, and clarify conditions of data re-use. Another solution for data-users to access trait data in a structured way is offered by FENNEC [@ankenbrand18], an online tool or self-hosted service to extract trait information from multiple sources for a target species community.  <!-- AG:  EDIT Platform for Cybertaxonomy (cybertaxonomy); NEON Dataportal http://data.neonscience.org/home;   -->

## Summing-up

Initiatives for standardisation (e.g. ontologies and data standards) and platforms for data management (e.g. database and data management platforms) provide great visibility, accessability and improved interoperability of datasets, but raise relatively high thresholds for data and metadata preparation. 
On the other end, public repositories on file-hosting services offer the publication of scientific primary data according to current funding and journal criteria, attracting a wealth of heterogeneous trait datasets. Due to the low thresholds concerning data standardisation, compilation of these distributed data sets is currently laborious. 
The goal must be to better integrate these distributed data into the global biodiversity data-management ecosystem by raising awareness for data standardisation, accessibility and re-useability on the side of data providers. By establishing this as a best practice in trait-based research, trait-data will eventually fullfill the FAIR guiding principles for scientific data [@wilkinson16]. 
In the following sections, we propose the development of vocabularies that lower the thresholds and offer high pay-off in the visibility and reuse of published data.

@@

As a response to the challenges outlined above, we propose a versatile vocabulary for trait-based ecological research. 
The standard vocabulary is accessible at https://ecologicaltraitdata.github.io/ETS/.
The aim of the vocabulary is to cover the variety of trait-based approaches and their different degrees of measurement detail. 
Rather than describing a data structure for relational databases, the vocabulary is intended as a more inclusive terminology, that can be used in three major use cases:

@@

By using this standard vocabulary, authors can ensure that the description of trait measurements that are uploaded to distributed data repositories will be unambiguous and generally applicable. It will facilitate re-use of data for future data aggregation initiatives and data synthesis and ensure long-term accessibility. By adopting the standard vocabulary, trait-data will fulfill the FAIR principles for scientific data [@wilkinson16]. 
We describe best practice principles and give examples for the application of the vocabulary in Supplementary Material B.

In designing this vocabulary, we drew on the combined expertise of empirical biodiversity researchers (data providers), biodiversity synthesis researchers (data users), and biodiversity informatics researchers (data managers).
We paid particular consideration to existing data standards (particularly @kattge11; @kattge11a; @garnier17; @parr16) to ensure compatibility of our proposed data structure with major trait databases and existing standards for biodiversity data management. 
<!--Besides being used for the publication of datasets, the standard vocabulary could be imposed in webservices or download tools, e.g. APIs that provide direct access to online databases. -->
<!--
## Why use the vocabulary

@@

The core terms describe a dataset according to the Entity-Quality model described above [@garnier17]: each entry describes a trait value (i.e. quality) observed on an individual or population (i.e. entity), of a biological taxon.
<!--Beyond these core observations, further information might be available that are related to the taxonomic assignment, or that put the reported fact, measurement or sampling event in a broader observation context (including geolocation and date information). 
These information may be useful for future analysis of the causal reasons of trait variation and should always be published along with the core data. For this case, -->
We offer three extensions of the core vocabulary ("Taxon", "Measurement or Fact", and "Occurrence"), which may simply be added as extra columns to the core dataset (see sections below for extensions and best practice when publishing primary data).
Additional terms are provided for metadata and for relating trait names to definitions and external ontologies or thesauri (see section on metadata below).

Some of the terms expand and refine terms of the Darwin Core and its Extensions (http://rs.tdwg.org/dwc/terms/) or the Glossary of EOL (http://eol.org/info/516). Wherever our vocabular refines or duplicates existing terms, we indicate this in the fields 'refines' or 'identical', respectively.
Any further columns on co-variates or measurement context, or other dataset-specific content, might draw from terms provided by Darwin Core and it's extensions, or be added as user-defined columns. However, these should be clearly specified in the meta-information accompanying the dataset. 
The scope of the vocabulary may not yet cover all aspects of morphological and evolutionary perspectives. 
Also, information about interactions between species are not within the scope of the Entity-Quality Model, but may easily be combined with trait data by using other extensions of DwC. 
The vocabulary proposed is intended to form the base of a standard exchange format.
Therefore, we invite researchers to contribute to the next iterations of the standard vocabulary and develop own applications and ontologies that interact with it.

## Specification of core terms

@@

## Extensions for additional data layers

@@

Data publications of traits should apply the FAIR principles of scientific data management and aim at findability, accessibility, interoperability and reusability [@wilkinson16]. By drawing column names and applying definitions of the ecological trait-data standard vocabulary, and providing reference to this standard in the metadata, authors can ensure that these principles are followed.

Above all, when publishing primary data, priority is on containing measurement values while minimizing the risk of purporting errors through data conversion or aggregation. Thus, data should be published as they have been recorded, i.e. in original units and raw values, along with standardised values that match a unified and compatible scheme of a data standard. This is true for the core terms, i.e. the trait value (the quality) and taxon name (the entity), but also for any co-variates reported along with it. 
For the core terms, data providers should use the '\*Std'-terms for standardised entries. For co-variates, the extensions of the vocabulary provide column names for standardised values only, e.g. elevation above sea-level. If any conversion was applied to obtain these data, users should keep the original values in user-defined columns, e.g. by appending '\*Orig' to the column name. Also, additional co-variates, e.g. on environmental growth conditions of the specimen, should be provided in user-defined columns. Any user-defined columns must be adequately defined in the accompanying meta-information.

When applying the vocabulary in a primary data publication, most authors will chose a two-dimensional spreadsheet rather than relational databases for reasons of simplicity. 
For spreadsheets, it is recommended to use a observation long-table format for the data (Fig. 2 *b*), rather than a species$\times$traits matrix (Fig. 2 *a*). 
As the long-table format draws from a defined set of columns, merging datasets is easier. Long-table datasets also purport multiple advantages for data manipulation [e.g. filtering, sub-setting and aggregating data, @wickham14].
